### 2018-06-29 contents/abstract.tex
final submission

\begin{abstract}
\addcontentsline{toc}{chapter}{Abstract}
Idiopathic (Classic) Trigeminal Neuralgia (TN) is a facial neuropathic pain syndrome characterized by paroxysmal, shock-like pain condition affecting one or more of the three trigeminal nerve (CNV) branches. TN is believed to be associated with nerve-vascular compression in the CNV root-entry-zone, but its pathophysiology is still unclear. Single-tensor diffusion tensor neuroimaging (DTI) studies of CN V revealed diffusivity changes in the cistern segment. However, the portion of the nerve within the brainstem remained elusive due to the limits of DTI, and the lack of reliable software to researchers. Thereby the study of TN using diffusion imaging is time-consuming and error-prone due to the complete manual involvement in all steps of data process, placement of region-of-interest (ROI), and tractography parameter tuning, and data analysis. 
There are two aims of this thesis: 1) to develop a fully-automated software framework to collate and analyze diffusion tractography to reduce human error and increase iteration speed of experiments. 2) to apply the method developed to the analysis of the trigeminal system in a more substantial patient group, and apply state-of-the-art diffusion imaging methods to advance the study of TN further. 
The specific aims are: a) establish the feasibility of high angular resolution diffusion tractography of brainstem CNV; b) establish the best way to minimize DWI to T1 co-registration error allow non-linear registration of tractography across multiple subjects; c) Create the software framework to generate and quantify tractography at the group level; d) apply the methodology to the reconstruction and quantification of the trigeminal sensory pathway. 
Towards these goals, in Study I, we establish the feasibility of applying spherical deconvolution tractography to delineate the full course of CNV from the trigeminal nucleus and demonstrate that TN is uniquely identified by disruptions in the cistern/REZ, while MS-TN can be distinguished by disruptions in the brainstem course of the nerve. In Study II, we determine that the best T1-DWI co-registration scalar is the Mean DWI image, and that the registration can potentially be improved with a novel spherical harmonic parameter scalar imaged called Anisotropic Power image. In Study III, we present the Selective Automated Group Integrated Tractography (SAGIT) software framework. Finally, in Study IV, we deploy end-to-end machine-learning TN classification to auto-discovers diffusivity disruptions in the cistern/REZ CNV, the trigeminopontothalamic decussaion, and thalamocortico S1 pathway. 
In sum, this thesis presents a detailed road-map of both the development and application of a novel end-to-end diffusion tractography machine learning classification. The application of this system to TN revealed specific diffusivity changes in all three segments of the trigeminal CN V, pontine decussation, and S1  white matter pathways, and pin-points the exact location of the diffusivity disruption at millimetre level.
\end{abstract}

>>>>

\begin{abstract}
\addcontentsline{toc}{chapter}{Abstract}
Idiopathic (Classic) Trigeminal Neuralgia (TN) is a facial neuropathic pain syndrome characterized by paroxysmal, shock-like pain condition affecting one or more of the three trigeminal nerve (CNV) branches. TN is believed to be associated with nerve-vascular compression in the CNV root-entry-zone, but its pathophysiology is still unclear. Single-tensor diffusion tensor neuroimaging (DTI) studies of CN V revealed diffusivity changes in the cistern segment. However, the portion of the nerve within the brainstem remained elusive due to the limits of DTI, and the lack of reliable software to researchers. Thereby the study of TN using diffusion imaging is time-consuming and error-prone due to the complete manual involvement in all steps of data process, placement of region-of-interest (ROI), and tractography parameter tuning, and data analysis. 
There are two aims of this thesis: 1) to develop a fully-automated software framework to collate and analyze diffusion tractography to reduce human error and increase iteration speed of experiments. 2) to apply the method developed to the analysis of the trigeminal system in a more substantial patient group, and apply state-of-the-art diffusion imaging methods to advance the study of TN further. 
The specific aims are: a) establish the feasibility of high angular resolution diffusion tractography of brainstem CNV; b) establish the best way to minimize DWI to T1 co-registration error allow non-linear registration of tractography across multiple subjects; c) Create the software framework to generate and quantify tractography at the group level; d) apply the methodology to the reconstruction and quantification of the trigeminal sensory pathway. 
Towards these goals, in Study I, we establish the feasibility of applying spherical deconvolution tractography to delineate the full course of CNV from the trigeminal nucleus and demonstrate that TN is uniquely identified by disruptions in the cistern/REZ, while MS-TN can be distinguished by disruptions in the brainstem course of the nerve. In Study II, we determine that the best T1-DWI co-registration scalar is the Mean DWI image, and that the registration can potentially be improved with a novel spherical harmonic parameter scalar imaged called Anisotropic Power image. In Study III, we present the Selective Automated Group Integrated Tractography (SAGIT) processing pipeline framework. Finally, in Study IV, we deploy end-to-end machine-learning TN classification to automatically discover diffusivity disruptions in the cistern/REZ CNV, the trigeminopontothalamic decussaion, and thalamocortico S1 pathway. 
In sum, this thesis presents a detailed road-map of both the development and application of a novel end-to-end diffusion tractography machine learning classification. The application of this system to TN revealed specific diffusivity changes in all three segments of the trigeminal CN V, pontine decussation, and S1  white matter pathways, and pin-points the exact location of the diffusivity disruption at millimetre level.
\end{abstract}

### 2018-06-29 contents/ap_paper.tex
Updates from ShareLaTeX
final submission

\begin{figure}[p]
\includegraphics[width=\linewidth]{figure4}
\caption[Visual comparisons of MDWI to T1 co-registrations.]{Visual comparisons of MDWI to T1 co-registrations. 
Shown are examples of MDWI images from one subject deformed using different registration transforms derived from the different image intermediates. 
T1 is overlaid as thresholded outlines in red. 
Primary differences in registration quality can be found in the brainstem (left and middle columns), and in the lateral ventricles, especially in the ventricular-caudate boundary (right column). }
\centering
\label{fig:APfigure4}
\end{figure}

@@

        \begin{figure}[p]
        \includegraphics[width=\linewidth]{figure6}
        \caption[Percent changes of similarity score from \protect $T_{\text{Affine}}$+SyN to \protect $T_{\text{Identity}}$ with CC as cost function.]{
        Percent changes of similarity score from \protect $T_{\text{Affine}}$+SyN to \protect $T_{\text{Identity}}$ with CC as cost function. Top row shows the violin plot distribution of CC and MI percent change under All Deformations and Autodeformations. Bottom row plots the statistically significant pair-wise correlations of the different transforms. 
        Under All Deformations (\protect $D_{\text{all}}$), CC percent change of FA is shown to be significantly lower (p<0.05) than that of all other transform types. GFA can be observed to have higher percent change than that of B0, MDWI, and FA. Under Autodeformations (\protect $D_{\text{auto}}$), CC percent change of FA and GFA are significantly higher than others, while AP has lower percent change. MI shows little percentage between all the images; GFA shows significantly differences under \protect $D_{\text{auto}}$, but also shows high variability.}
        \centering
        \label{fig:APfigure6}
        \end{figure}
    \end{landscape}

>>>>

\begin{figure}[p]
\centering
\includegraphics[width=0.9\linewidth]{figure4}
\caption[Visual comparisons of MDWI to T1 co-registrations.]{Visual comparisons of MDWI to T1 co-registrations. 
Shown are examples of MDWI images from one subject deformed using different registration transforms derived from the different image intermediates. 
T1 is overlaid as thresholded outlines in red. 
Primary differences in registration quality can be found in the brainstem (left and middle columns), and in the lateral ventricles, especially in the ventricular-caudate boundary (right column). }
\label{fig:APfigure4}
\end{figure}

@@

        \begin{figure}[p]
        \includegraphics[width=\linewidth]{figure6}
        \caption[Percent changes of similarity score from \protect $T_{\text{Affine}}$+SyN to \protect $T_{\text{Identity}}$ with CC as cost function.]{
        Percent changes of similarity score from \protect $T_{\text{Affine}}$+SyN to \protect $T_{\text{Identity}}$ with CC as cost function. Top row shows the violin plot distribution of CC and MI percent change under All Deformations and Autodeformations. Bottom row plots the statistically significant pair-wise correlations of the different transforms. 
        Under All Deformations (\protect $D_{\text{all}}$), CC percent change of FA is shown to be significantly lower (p<0.05) than that of all other transform types. GFA can be observed to have higher percent change than that of B0, MDWI, and FA. Under Autodeformations (\protect $D_{\text{auto}}$), CC percent change of FA and GFA are significantly higher than others, while AP has lower percent change. MI shows little percentage changes between all the images; GFA shows significantly differences under \protect $D_{\text{auto}}$, but also shows high variability.}
        \centering
        \label{fig:APfigure6}
        \end{figure}
    \end{landscape}

### 2018-06-29 contents/reviews/dmri.tex
Updates from ShareLaTeX
final submission

Diffusion magnetic resonance imaging (dMRI) is a popular MR modality that is capable of providing orientation information of brain white matter fibres \textit{in-vivo}, and thereby permitting the reconstruction and analysis of brain white matter structures with the use of computer reconstruction techniques.

@@

\begin{figure}[ht]
\begin{center}
\begin{tabular}{c}
\includegraphics[width=\textwidth]{csd-angles}
\end{tabular}
\caption{ Comparison of angle discrimination ability and CSD FOD. B-value=3000 \protect $s/mm^2$. Taken from \protect \cite{Tournier2007b} } 
\label{fig:csd}
\end{center}
\end{figure}

@@

Tractography is a computer graphics technique used by neuroscientists to delineate neural pathways with DWI imaging. Applying pathway tracking to DWI imaging in the brain was first explored by Mori \cite{Mori2002b} and Basser\cite{Basser2000}. Tractography in essence is the construction, or propagation of streamline from an underlying field of orientation information, in order to visualize continuous structures within a volume.  The orientation field is generally provided as a 3D matrix of directional information obtained from DTI, CSD, or other methods. For example, in DTI the directions are the eigen vector with the greatest magnitude $\hat{\lambda_1}$; for CSD, the directional vectors would be peaks from the FOD.

>>>>

Diffusion magnetic resonance imaging (dMRI) is a popular MR modality that is capable of providing orientation information of brain white matter fibres \textit{in-vivo}, and thereby permitting the reconstruction and analysis of brain white matter structures with the use of computer reconstruction techniques \cite{Johansen-Berg2013}.

@@

\begin{figure}[ht]
\begin{center}
\begin{tabular}{c}
\includegraphics[width=\textwidth]{csd-angles}
\end{tabular}
\caption{ Comparison of angle discrimination ability and CSD FOD. B-value=3000 \protect $s/mm^2$. Taken from \protect \cite{Tournier2007b}. Reproduced with permission from Copyright Clearance Center (License ID: 4316770133531) } 
\label{fig:csd}
\end{center}
\end{figure}

@@

Tractography is a computer graphics technique used by neuroscientists to delineate neural pathways with DWI imaging. Applying pathway tracking to DWI imaging in the brain was first explored by Mori \cite{Mori2002b} and Basser\cite{Basser2000}. Tractography in essence is the construction, or propagation of streamlines from an underlying field of orientation information, in order to visualize continuous structures within a volume.  The orientation field is generally provided as a 3D matrix of directional information obtained from DTI, CSD, or other methods. For example, in DTI the directions are the eigen vector with the greatest magnitude $\hat{\lambda_1}$; for CSD, the directional vectors would be peaks from the FOD.

### 2018-06-29 contents/reviews/machine-learning.tex
final submission

Increasing complexity in neuroimaging data has led to widespread applications of multi-variable machine learning methods in neuroimaging. Application of ML in pain neuroimaging became controversial when Tor Wager \cite{Wager2013} demonstrated a Lasso-PCA model that classified acute experimental pain with sensitivity and specificity of up to 90\% solely from fMRI imaging data. Additional studies showed consistently good performance of ML-based nociceptive pain discrimination. Discrimination of chronic pain from nociceptive pain proved to be more challenging, due to the high degree of similarity in activation patterns. The development of pain detection through brain imaging led to debates of the validity of presenting such methods as evidences in courtroom, where Davis et al argued against such actions \cite{Davis2012a}. While it is commonly accepted that pain evokes activity in the brain in multiple regions of the dynamic pain connectome, some studies suggest that the dorsal posterior-insula is heavily involved in chronic pain. Segerdale et al claimed that the dorsal posterior insula is a primary pain region \cite{Segerdahl2015a}, and this proposition ignited renewed debates surrounding regional specific brain centers for chronic pain \cite{Davis2015}.

>>>>

Increasing complexity in neuroimaging data has led to widespread applications of multi-variable machine learning methods in neuroimaging. Application of ML in pain neuroimaging became controversial when Tor Wager \cite{Wager2013} demonstrated a Lasso-PCA model that classified acute experimental pain with sensitivity and specificity of up to 90\% solely from fMRI imaging data. Additional studies showed consistently good performance of ML-based nociceptive pain discrimination. Detecting chronic pain proved to be more challenging, due to the high degree of similarity between acute and chronic pain fMRI activation patterns. The development of pain detection through brain imaging led to debates of the validity of presenting such methods as evidences in courtroom, where Davis et al argued against such actions \cite{Davis2012a,Davis2017}. While it is commonly accepted that pain evokes activity in the brain in multiple regions of the dynamic pain connectome, some studies suggest that the dorsal posterior-insula is heavily involved in chronic pain. Segerdale et al claimed that the dorsal posterior insula is a primary pain region \cite{Segerdahl2015a}, and this proposition ignited renewed debates surrounding regional specific brain centers for chronic pain \cite{Davis2015}.

### 2018-06-29 contents/reviews/mri.tex
Updates from ShareLaTeX
final submission

MRI is based on the principles of nuclear magnetic resonance (NMR), where atomic nuclei, when placed in a magnetic field, can absorb and emit electromagnetic radiation. An MRI machine applies a primary magnetic field (B0) to magnetize and align the hydrogen nuclei in the body. A radio frequency (RF) field is subsequently introduced perpendicular to B0, and reorients the spin axis. This causes the nuclei to produce NMR signals detectable by the MRI scanner. Based on the precession frequency decay, the signals are recorded to construct an image of the scanned area based on the contrasts in water and fat concentrations.

@@

\begin{figure}[htb]
\begin{center}
\begin{tabular}{c}
\includegraphics[width=\textwidth]{mri/rf-pulse.pdf}
\end{tabular}
\caption{Illustration of the process of transverse relaxation. Initially, the spin precesses about the B0 field under steady-state. When a 90deg RF pulse is applied, the spin orients towards the xy-plane. When the RF pulse is switched off, the spin now relaxes towards the z-axis.  } \label{fig:relaxation}
\end{center}
\end{figure}

@@

\begin{figure}[htb]
\begin{center}
\begin{tabular}{c}
\includegraphics[width=\textwidth]{mri/spin-echo.pdf}
\end{tabular}
\caption{Illustration of the process of spin-echo. Initially, the spins orients to wards B0 field under steady-state. When a 90deg RF pulse is applied, the spins orients to the xy-plane. When the RF pulse is switched off, the spins undergoe phase decoherence. A 180 degrees refocusing pulse is applied to bring spins back into coherence, at which point in time the peak of an "echo" occurs.  } \label{fig:spin-echo}
\end{center}
\end{figure}

>>>>

MRI is based on the principles of nuclear magnetic resonance (NMR), where atomic nuclei, when placed in a magnetic field, can absorb and emit electromagnetic radiation. An MRI machine applies a primary magnetic field (B0) to magnetize and align the hydrogen nuclei in the body. A radio frequency (RF) field is subsequently introduced perpendicular to B0, and reorients the spin axis. This causes the nuclei to produce NMR signals detectable by the MRI scanner. Based on the precession frequency decay, the signals are recorded to construct an image of the scanned area based on the contrasts in water and fat concentrations \cite{Plewes2012,Johansen-Berg2013}.

@@

\begin{figure}[htb]
\begin{center}
\begin{tabular}{c}
\includegraphics[width=\textwidth]{mri/rf-pulse.pdf}
\end{tabular}
\caption{Illustration of the process of transverse relaxation. Initially, the net magnetization orientes to the B0 field under steady-state. When a 90 deg RF pulse is applied, the spin orients towards the xy-plane. When the RF pulse is switched off, the spin now relaxes towards the z-axis.  } \label{fig:relaxation}
\end{center}
\end{figure}

@@

\begin{figure}[htb]
\begin{center}
\begin{tabular}{c}
\includegraphics[width=\textwidth]{mri/spin-echo-2.pdf}
\end{tabular}
\caption{Illustration of the process of spin-echo. Initially, the spins orients to wards B0 field under steady-state. When a 90deg RF pulse is applied, the spins orients to the xy-plane. When the RF pulse is switched off, the spins undergoe phase decoherence. A 180 degrees refocusing pulse is applied to bring spins back into coherence, at which point in time the peak of an "echo" occurs.  } \label{fig:spin-echo}
\end{center}
\end{figure}

### 2018-06-29 contents/reviews/multiplesclerosis.tex
final submission

\subsection{Pain and MS}
Pain is a common co-morbidity of MS, with up to 50-85\% prevalence \cite{Osterberg2005}. Common pain symptoms include acute pain such as optic neuritis and painful tonic seizure, as well as chronic pain such as TN. 91\% of the patients report pain at the time of evaluation, and most experience daily pain. Both central and peripheral pain are reports in people of MS. Central pain in MS is reported to occur 27-58\% of the subjects \cite{OConnor2008}.

The most common pain in MS includes extremity pain and TN (MS-TN), which is also termed symptomatic TN. MS-TN has a prevalence 2\% in the MS, which is 20 times higher than in a general population \cite{Cruccu2009}. The clinical presentation of the pain is similar to classic TN. MS-TN has a prevalence rate of about 2-4\%, and in some studies, there are reports of prevalence rate as high as 6\%. Unilateral MS-TN is the most common pain type. However, there are up to 30\% occurrences of bilateral pain. MS-TN pathophysiology is believed to be different from classic TN, as it is often accompanied by the presentation of pontine MS lesions \cite{Cruccu2009}. The treatment of MS-TN via ablative interventions is also less effective than in classic TN \cite{Nurmikko2009}. The exact pathophysiological differences between MS-TN and classic TN, however, are not known. While it shares the feature of vascular compression with classic TN, its relevance to MS-TN pain has not been precisely determined \cite{Nurmikko2009}.

>>>>

\subsection{Pain and MS}
Pain is a common co-morbidity of MS, with up to 50-85\% prevalence \cite{Osterberg2005}. Common pain symptoms include acute pain such as optic neuritis and painful tonic seizure, as well as chronic pain such as TN. 91\% of the patients report pain at the time of evaluation, and most experience daily pain, where both central and peripheral pain are reported in people of MS. Central pain in MS is reported to occur 27-58\% of the subjects \cite{OConnor2008}.

The most common pain in MS includes extremity pain and TN (MS-TN), which is also termed symptomatic TN. MS-TN has a prevalence 2\% in the MS population, which is 20 times higher than in a general population \cite{Cruccu2009}. The clinical presentation of the pain is similar to classic TN. MS-TN has a prevalence rate of about 2-4\%, and in some studies, there are reports of prevalence rate as high as 6\%. Unilateral MS-TN is the most common pain type. However, there are up to 30\% occurrences of bilateral pain. MS-TN pathophysiology is believed to be different from classic TN, as it is often accompanied by the presentation of pontine MS lesions \cite{Cruccu2009}. The treatment of MS-TN via ablative interventions is also less effective than in classic TN \cite{Nurmikko2009}. The exact pathophysiological differences between MS-TN and classic TN, however, are not known. While it shares the feature of vascular compression with classic TN, its relevance to MS-TN pain has not been precisely determined \cite{Nurmikko2009}.
